---
output: 
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
geometry: margin=2.54cm
title: "Blue Mesa Power Production Analysis"
subtitle: "Web address for GitHub repository"
author: "Stephanie Kinser, Katie Owens, Cassidy White"
fontsize: 12pt
mainfont: Times New Roman
editor_options: 
  chunk_output_type: console
---

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory
getwd()

# Load your packages
#install.packages("dplyr")
library(dplyr)
library(tidyverse)
library(ggplot2)
library(lubridate)
#install.packages("tidyr")
library(tidyr)
#install.packages("zoo")
library(zoo)
#install.packages("agricolae")
library(agricolae)
#install.packages("corrplot")
library(corrplot)
#install.packages("splitstackshape")
library(splitstackshape)
#install.packages("matrixStats")
library(matrixStats)

# Set your ggplot theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right") + 
  theme_bw()
theme_set(mytheme)

# Load your datasets
climate<-read.csv("./Data/Processed/BlueMesaClimate.csv")
power<-read.csv("./Data/Processed/BlueMesaPower.csv")
reservoir<-read.csv("./Data/Processed/BlueMesaReservoir.csv")

```


# Rationale and Research Questions

Weather patterns have a significant impact on business, health, and community well-being. The western United States has been experiencing years of low precipitation and prolonged drought.  Currently, all of Colorado is experience a moderate to extreme drought (Drought.gov).  The National Oceanic and Atmospheric Administration's (NOAA) National Integrated Drought Information System (NIDIS) tracks the impacts of drought, including its effect on crops and livestock and fire.  Across Colorado, the impact of drought is clear in the low levels of reservoirs.  

 


\newpage

# Dataset Information



\newpage

# Exploratory Analysis 

```{r summary stats, echo=FALSE}
#read in data
all.data<-read.csv("../Data/Processed/AllData.csv")
all.data$Date<-as.Date(all.data$Date, format = "%Y-%d-%m")
all.data<-all.data %>%
  na.omit()
all.data.nodate<-select(all.data, -c(X, Date))

#adjust data to remove Spillway and rename columns
all.data.nospill <-
  select(all.data.nodate, -c(Spillway.cfs)) %>%
  rename("Max Temperature" = MaxT,
         "Min Temperature" = MinT,
         "Precipitation" = Precip,
         "Snowfall" = Snow,
         "Electricity Generation.MWh" = MWH)

#summary statistics table data
summary.table<-all.data.nospill %>%
  summary()
summary.table<-as.data.frame(summary.table) %>% cSplit("Freq", sep = ":", type.convert = FALSE) 
summary.table<-summary.table %>%
  select(Var2, Freq_1, Freq_2)
summary.table<-pivot_wider(summary.table, names_from = Freq_1, values_from = Freq_2)
summary.table<-select(summary.table, c(Var2, Mean, Min., Max.))

sd<-colSds(as.matrix(all.data.nospill[sapply(all.data.nospill, is.numeric)]))
sd<-as.data.frame(sd)
summary.table<-cbind(summary.table, sd)

names(summary.table)[names(summary.table)=='Var2']<-'Variable'

#change data to numeric
summary.table$Mean <- as.numeric(summary.table$Mean)
summary.table$Max. <- as.numeric(summary.table$Max.)
summary.table$Min. <- as.numeric(summary.table$Min.)
summary.table$sd <- as.numeric(summary.table$sd)

#create gt table
gt(summary.table) %>%
    tab_header(title = "Blue Mesa Reservoir Summary Statistics") %>%
    fmt_number(
      columns = c(Mean, Min., Max., sd), decimals = 2)

```


```{r wrangle the data, include = FALSE}
#getting dates in the proper format 
climate$Date <-as.Date(climate$Date, format = "%Y-%m-%d")

power$Date <- as.Date(power$Date, format = "%Y-%m-%d")

reservoir$Date <- as.Date(reservoir$Date, format = "%d-%b-%y")

#make a gathered data set for data viz
climate.longer<-pivot_longer(climate, MaxT:MinT, names_to = "Min/Max", values_to = "Temperature")

#make a gathered data set for data viz
reservoir.longer<-reservoir %>%
  pivot_longer(Inflow.cfs:Total.cfs, names_to = "Flow_Type", values_to = "Flow_(cfs)") %>%
  pivot_longer(Storage.af:Evaporation.af, names_to = "Volume_Type", values_to = "Volume_(af)")

#combine power and reservoir data
all.data<-left_join(reservoir, climate)
all.data<-left_join(all.data, power)
#write.csv(all.data, "AllData.csv") #commenting out b/c already written
```


```{r visual exploration}
#min and max temperature over time
ggplot(climate.longer) +
  geom_point(aes(x = Date, y = Temperature, group = `Min/Max`, color = `Min/Max`)) +
  labs(y="Temperature (F)", color = "Legend")+
  scale_color_manual(labels=c("Maximum", "Minimum"), values=c("red", "blue"))

#power production over time
ggplot(power, aes(x=Date, y=MWH))+
  geom_line()+
  ylab("Electricity Generation (MWh)")

#reservoir elevation over time
ggplot(reservoir, aes(x=Date, y = Elevation.ft))+
  geom_line(color = "blue")+
  ylab("Reservoir Elevation (ft)")

#reservoir volume over time
ggplot(reservoir.longer, aes(x=Date, y = `Volume_(af)`, color = `Volume_Type`))+
  geom_line()+
  scale_y_continuous(labels = scales::comma)+
  labs(y = "Volume (af)", color = "Legend")+
  scale_color_manual(labels=c("Evaporation", "Storage"), values=c("orange", "blue"))

#reservoir total flow over time
ggplot(reservoir, aes(x=Date, y = Total.cfs))+
  geom_line(color = "blue")+
  labs(y = "Total Flow (cfs)")

#reservoir power production by power.cfs
#I want to do a second axis here but it isn't working. Would also need to add a legend.  SK: I made the second axis appear but it's not adjusting to the right scale.
ggplot(all.data, aes(x=Date))+
  geom_line(aes(y=MWH), color = "red")+
  geom_line(aes(y=Power.cfs), color = "blue")+
  #scale_color_manual(name = "Legend", labels = c("Power Flow (cfs)", "Electricity Generation (MWh)"), values = c("blue", "red"))+
  scale_y_continuous(
    name = "Electricity Generation (MWh)",
    sec.axis = sec_axis(~./100, name ="Power Flow (cfs)"))
  
```

\newpage

# Analysis



## Question 1: <insert specific question here and add additional subsections for additional questions below, if needed>

## Question 2: 




\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 
