---
output: 
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
geometry: margin=2.54cm
title: "Blue Mesa Power Production Analysis"
subtitle: "https://github.com/stephk27/KinserOwensWhite_ENV872_EDA_FinalProject"
author: "Stephanie Kinser, Katie Owens, Cassidy White"
fontsize: 12pt
mainfont: Times New Roman
editor_options: 
  chunk_output_type: console
---

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory
getwd()

# Load your packages
#install.packages("dplyr")
library(dplyr)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(gt)
#install.packages("tidyr")
library(tidyr)
#install.packages("zoo")
library(zoo)
#install.packages("agricolae")
library(agricolae)
#install.packages("corrplot")
library(corrplot)
#install.packages("splitstackshape")
library(splitstackshape)
#install.packages("matrixStats")
library(matrixStats)
#install.packages("gtsummary")
library(gtsummary)
#install.packages("knitr")
require(knitr)
#install.packages("kfigr")
require(kfigr)

# Set your ggplot theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right") + 
  theme_bw()
theme_set(mytheme)

# Load your datasets
climate<-read.csv("../Data/Processed/BlueMesaClimate.csv")
power<-read.csv("../Data/Processed/BlueMesaPower.csv")
reservoir<-read.csv("../Data/Processed/BlueMesaReservoir.csv")

```


# Rationale and Research Questions

Climate patterns have a significant impact on business, health, and community well-being. The western United States has been experiencing years of low precipitation and prolonged drought.  Currently, all of Colorado is experiencing a moderate to extreme drought (Drought.gov).  The National Oceanic and Atmospheric Administration's (NOAA) National Integrated Drought Information System (NIDIS) tracks the impacts of drought, including its effect on crops and livestock and fire.  We want to explore the effects of climate patterns on hydroelectric power production, an impact that is not currently captured by NIDIS.
\
Hydroelectric power plants rely on a sufficient reservoir level to maintain power output, which can be an important source of carbon-free electricity.  We wonder how climate patterns and reservoir management have impacted hydroelectric power production.  We identified Blue Mesa Reservoir and Hydroelectric Power Plant in Colorado to explore our research questions.
\
Blue Mesa Power Plant has a nameplate capacity of 86.2 MW and began generating power in 1967.  It is located on the Gunnison River in Colorado and it is managed by the Colorado Bureau of Reclamation.  The reservoir also provides long-term storage to ensure that Colorado is able to meet its downstream flow obligation as established in the Colorado River Compact (CO Bureau of Reclamation).
\
**Research Questions:** \

-Which climate and reservoir management variables influence Blue Mesa's power production?
-How has electricity generation of Blue Mesa Power Plant changed over time?


\newpage

# Dataset Information

For our analysis, we relied on two data sets.  For Blue Mesa's power production and reservoir management, we sourced data from the Colorado Bureau of Reclamation.  The power production dataset includes monthly electricity generation reported in MWh.  The reservoir management dataset includes eight variables, including elevation, storage, inflow, and outflow.  We pulled monthly averages for this data from 2003 to 2021 when all the data was available.
\
We sourced climate data from the Colorado Climate Center, an institute at Colorado State University.  We pulled monthly data for Blue Mesa on the maximum temperature, minimum temperature, average precipitation, and average snowfall.


\newpage

# Exploratory Analysis 



```{r wrangle data part one, include=FALSE}
#read in all data
all.data<-read.csv("../Data/Processed/AllData.csv")
all.data$Date<-as.Date(all.data$Date, format = "%Y-%d-%m")
all.data<-all.data %>%
  na.omit()
all.data.nodate<-select(all.data, -c(X, Date))%>%
  rename("Max Temperature" = MaxT,
         "Min Temperature" = MinT,
         "Precipitation" = Precip,
         "Snowfall" = Snow,
         "Electricity Generation.MWh" = MWH)

```

Prior to analysis, data were explored in terms of both summary statistics and visual patterns and relationships. Table \ref{table: summary_statistics} summarizes the data for all fourteen variables used in building the explanatory model. Summary statistics include mean, minimum value, maximum value, and standard deviation across all observations. From 2003 to 2021, Blue Mesa Reservoir's elevation ranged from `r round(min(all.data$Elevation.ft), 0)` ft to `r round(max(all.data$Elevation.ft), 0)`. Temperature at the reservoir ranged from -`r round(min(all.data$MinT), 1)` $^\circ$F to `r round(max(all.data$MaxT), 1)` $^\circ$F. The Blue Mesa dam produced an average of `r round(mean(all.data$MWH), 0)` MWh of electricity per month with a range of `r round(min(all.data$MWH), 0)` MWh to `r round(max(all.data$MWH), 0)` MWh during peak production.

```{r build summary statistics table inputs, echo=FALSE}

#summary statistics table data
summary.table<-all.data.nodate %>%
  summary()
summary.table<-as.data.frame(summary.table) %>% cSplit("Freq", sep = ":", type.convert = FALSE)
summary.table<-summary.table %>%
   select(Var2, Freq_1, Freq_2)
summary.table<-pivot_wider(summary.table, names_from = Freq_1, values_from = Freq_2)
summary.table<-select(summary.table, c(Var2, Mean, Min., Max.))
 
sd<-colSds(as.matrix(all.data.nodate[sapply(all.data.nodate, is.numeric)]))
sd<-as.data.frame(sd)
summary.table<-cbind(summary.table, sd)

names(summary.table)[names(summary.table)=='Var2']<-'Variable'

#change data to numeric
summary.table$Mean <- as.numeric(summary.table$Mean)
summary.table$Max. <- as.numeric(summary.table$Max.)
summary.table$Min. <- as.numeric(summary.table$Min.)
summary.table$sd <- as.numeric(summary.table$sd)
```

```{r summary_statistics, echo=FALSE}
#create gt table
gt(summary.table) %>%
  tab_header(title = "Table 1: Blue Mesa Reservoir Summary Statistics")%>%
  fmt_number(columns = c(Mean, Min., Max., sd), decimals = 2)%>%
  cols_label(
    Min. = "Minimum",
    Max. = "Maximum",
    sd = "Standard Deviation"
  )
```

```{r wrangle the data part two, include = FALSE}
#getting dates in the proper format 
climate$Date <-as.Date(climate$Date, format = "%Y-%m-%d")

power$Date <- as.Date(power$Date, format = "%Y-%m-%d")

reservoir$Date <- as.Date(reservoir$Date, format = "%d-%b-%y")

#make a gathered data set for data viz
climate.longer<-pivot_longer(climate, MaxT:MinT, names_to = "Min/Max", values_to = "Temperature")

#make a gathered data set for data viz
reservoir.longer<-reservoir %>%
  pivot_longer(Inflow.cfs:Total.cfs, names_to = "Flow_Type", values_to = "Flow_(cfs)") %>%
  pivot_longer(Storage.af:Evaporation.af, names_to = "Volume_Type", values_to = "Volume_(af)") 
```

To further explore the data, a series of graphs were developed to visually examine trends in key variables over time as well as variables with expected trends in relation to each other. In `r figr('min/max', TRUE, type = "figure")`, both minimum and maximum temperatures are shown from 2003 until 2021 at Blue Mesa Reservoir. To examine trends in electricity generation over time, `r figr('power production', TRUE, type = "figure")` plots electricity production (MWh) by date. In the next plot, `r figr('elevation', TRUE, type = "figure")`, reservoir elevation (ft) is graphed over time followed by `r figr('volume', TRUE, type = "figure")` showing volume in acre-feet over time. Total inflow over time is then shown in `r figr('total inflow', TRUE, type = "figure")`.

```{r min/max, echo=FALSE, fig.align="center", anchor="figure", warning = FALSE, fig.cap= "Minimum and maximum temperature at Blue Mesa Reservoir in Fahrenheit from January 2003 until December 2021.", fig.height=3, fig.width=5}
#min and max temperature over time
ggplot(climate.longer) +
  geom_point(aes(x = Date, y = Temperature, group = `Min/Max`, color = `Min/Max`)) +
  labs(y="Temperature (F)", color = "Legend")+
  scale_color_manual(labels=c("Maximum", "Minimum"), values=c("red", "blue"))
```

```{r power production, echo=FALSE, fig.align="center", anchor="figure", fig.cap = "Power production in megawatt hours at Blue Mesa Reservoir from June 2003 until March 2021.", fig.height=3, fig.width=5}
#power production over time
ggplot(power, aes(x=Date, y=MWH))+
  geom_line()+
  labs(y="Electricity Generation (MWh)")+
  theme(plot.caption = element_text(hjust = .25))
```

```{r elevation, echo=FALSE, fig.align="center", anchor="figure", fig.cap = "Reservoir elevation in feet at Blue Mesa Reservoir from January 2003 until December 2021.", fig.height=3, fig.width=5}
#reservoir elevation over time
ggplot(reservoir, aes(x=Date, y = Elevation.ft))+
  geom_line(color = "blue")+
  labs(y = "Reservoir Elevation (ft)")
```

```{r volume, echo=FALSE, fig.align="center", anchor="figure", fig.cap = "Reservoir volume in acre-feet at Blue Mesa Reservoir from January 2003 until December 2021.", fig.height=3, fig.width=5}
#reservoir volume over time
ggplot(reservoir.longer, aes(x=Date, y = `Volume_(af)`, color = `Volume_Type`))+
  geom_line()+
  scale_y_continuous(labels = scales::comma)+
  labs(y = "Volume (af)", color = "Legend")+
  scale_color_manual(labels=c("Evaporation", "Storage"), values=c("orange", "blue"))
```

```{r total flow, echo=FALSE, fig.align="center", anchor="figure", fig.cap = "Total flow into Blue Mesa Reservoir in cubic feet per second from January 2003 until December 2021.", fig.height=3, fig.width=5}
#reservoir total flow over time
ggplot(reservoir, aes(x=Date, y = Total.cfs))+
  geom_line(color = "blue")+
  labs(y = "Total Inflow (cfs)")
```

From there, a few variables are explored graphically with respect to the dependent variable, electricity generation (MWh). In `r figr('MWHxTotal', TRUE, type = "figure")`, total inflow (cfs) is plotted against electricity generation and shows a positive relationship. As inflow increases, electricity generation increases as well. 

```{r MWHxTotal, echo=FALSE, fig.align="center", anchor="figure", fig.cap = "Relationship between total water inflow (cfs) and power production (MWH) at Blue Mesa Reservoir from 2003 until 2021.", fig.height=3, fig.width=5}
#MWH by total inflow
ggplot(all.data, aes(x=Total.cfs, y = MWH))+
  geom_line()+
  labs(x = "Total Inflow (cfs)", y = "Electricity Generation (MWh)") 
```

Power outflow is then plotted against electricity generation in `r figr('MWHxPower', TRUE, type = "figure")`, also showing a positive relationship. These trends were expected based on the high correlation between total inflow and electricity generation (MWh) and power outflow and electricity generation (MWh) as shown in `r figr('correlation', TRUE, type = "figure")`. However, the relationships between these variables are explored further in the analysis for significance.

```{r MWHxPower, echo=FALSE, fig.align="center", anchor="figure", fig.cap = "Relationship between power outflow (cfs) and power production (MWH) at Blue Mesa Reservoir from 2003 until 2021.", fig.height=3, fig.width=5}
#MWH by power flow
ggplot(all.data, aes(x=Power.cfs, y = MWH))+
  geom_line()+
  labs(x = "Power Outflow (cfs)", y = "Electricity Generation (MWh)")   
```

Last, `r figr('MWHxPrecip', TRUE, type = "figure")` shows precipitation (inches) plotted against electricity generation (MWh). No noticeable trend is observed but the contribution from precipitation in explaining variation in electricity generation is explored further in the analysis.

```{r MWHxPrecip, echo=FALSE, fig.align="center", anchor="figure", fig.cap = "Relationship between precipitation (inches) and power production (MWH) at Blue Mesa Reservoir from 2003 until 2021.", fig.height=3, fig.width=5}
#MWH by precipitation
ggplot(all.data, aes(x=Precip, y = MWH))+
  geom_line()+
  labs(x = "Precipitation (in)", y = "Electricity Generation (MWh)") 
```


\newpage

# Analysis

The analysis was divided into two components: a linear regression to explore the influence of other variables on electricity generation, and a time series analysis to explore how electricity generation has changed over time.

## Question 1: Which climate and reservoir management variables influence Blue Mesa's power production?

The first step in the analysis was to explore the correlation between different variables used in the model. In (`r figr('correlation', TRUE, type = "figure")`), narrow ovals or straight lines indicate two variables that are highly correlated. The wider the oval, the less correlated the variables are. This correlation matrix provides an understanding of how variables relate to one another and is useful when interpreting results.

```{r correlation, echo=FALSE, fig.align="center", anchor="figure", fig.cap = "Correlation matrix for all dependent and independent variables used in building the explanatory model.", fig.height=4, fig.width=6}
all.data.corr<-cor(all.data.nodate)
corrplot(all.data.corr, method = "ellipse")
```

A multiple linear regression was then performed to explore the contribution of the thirteen independent variables in explaining variation in the dependent variable, electricity generation (MWh). After the initial regression, an AIC test was run to identify which combination of variables contributed to the model of best fit for explaining electricity generation (MWh). Eight independent variables were selected for the final model and their coefficients (Beta) and significance values are summarized in Table \r{tab:Regression_Table}. 

```{r Multiple Linear Regression Analysis, include=FALSE}
#Run a regression with all variables to take a first look
regression.all<-lm(data = all.data, MWH ~ Elevation.ft + Storage.af + Evaporation.af + Inflow.cfs + UnregInflow.cfs + Bypass.cfs + Spillway.cfs +  MaxT + MinT + Precip + Snow + Total.cfs + Power.cfs)

summary(regression.all)

# #Run a stepwise AIC test to find the model of best fit from all the above variables
step(regression.all)

#Choose the model of best fit from the AIC test and run below
regression.final <- lm(data = all.data, MWH ~ Elevation.ft + Storage.af + Evaporation.af +
     Inflow.cfs + MaxT + Precip + Total.cfs + Power.cfs)

summary(regression.final)

r.squared<-summary(regression.final)$adj.r.squared
r.squared<-round(r.squared, digits = 3)

#write.csv(as.data.frame(summary(regression.final)$coef), file = "./Output/RegressionTable.csv")
```

```{r Regression_Table, echo=FALSE}
fit_stats <-
  broom::glance(regression.final) %>%
  select(`Adjusted R squared` = r.squared, AIC) %>%
  mutate_all(function(x) style_sigfig(x, digits = 3)) %>%
  {paste(names(.), ., sep = " = ", collapse = "; ")}

tbl_regression(regression.final) %>%
  as_gt() %>%
  tab_header(title = "Table 2: Regression Table")%>%
  gt::tab_source_note((gt::html(fit_stats)))
```

The final regression identifies elevation, storage, evaporation, inflow, maximum temperature, precipitation, total inflow, and power outflow as all contributing to explaining variation in electricity generation (MWh) across the study period. The model has an adjusted R^2 of `r r.squared`, indicating the model explains 99.5% of the variation in MWh. According to the model results, a decrease in elevation of 249 ft per month leads to an increase of 1 MWh of electricity generation. This result is expected considering that as water is drawn down through the hydroelectric turbines for electricity generation, the reservoir's elevation decreases. Likewise, as power outflow increases by 17 cfs, electricity generation increases by 1 MWh per month. This result also agrees with our expected findings since power outflow and electricity generation (MWh) are highly correlated and since power outflow is released with the intention of generating electricity.

## Question 2: How has electricity generation of Blue Mesa Power Plant changed over time?




\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 
